---
title: GSEA+
output: pdf_document
params:
  # Parameters for the plots
  add_title: NA
  description: NA
  title_size: NA
  title_position: NA
  color_panel: NA
  color_plot: NA
  color_title: NA
  color: NA
  color_dashed: NA
  curve_size: NA
  add_mes: NA
  add_0: NA
  add_panel2: NA
  color_panel2: NA
  color_search_gene: NA
  add_panel3: NA
  color_exp_line: NA
  color_exp: NA
  red_blue: NA
  line_exp: NA
  color_dashed1: NA
  
  # Data
  gene_list: NA
  candidate_list: NA
  num_prerank: NA
  num_geneset: NA
  names_prerank: NA
  names_geneset: NA
  selection_genes: NA
  
  # Tables
  important_values: NA
  important_statistics: NA
  
  nums_plots: num_prerank * num_geneset
  
  # pos_plots: NA
  # neg_plots: NA
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(ggplot2)
library(shiny) 
library(shinydashboard)
library(knitr) 
library(data.table)
library(rintrojs) 
library(shinyWidgets)
library(rmarkdown) 
library(cowplot) 
library(dplyr)
library(devtools)
library(gridExtra)
```


```{r functions, include=FALSE}
# Takes the names and values of the preranked list (1st and 2nd column)
# Returns the dataframe with the rank order
reading_inputs <- function(names, values){
  
  # Creating a data frame
  data <- data.frame(
    names,
    values
  )
  
  data <- data[order(data$values),] # Ordering the data frame by the value of expression
  
  ranking <- c(length(data$names):1)
  data["ranking"] <- ranking
  
  data <- data[order(data$ranking),] # Order the data frame by the ranking order
  
  return(data)
} # Step 0

# Takes the dataframe and the gene set
# Returns the updated dataframe and the sum of the overlapping genes
assessing_overlaping <- function(data, candidate_list){
  
  indicator_function <- c() 
  sum_in_set = 0 # Keep track of the total rank metric for the genes overlapping
  
  for(i in data$ranking){ 
    if(data$names[i] %in% candidate_list){
      indicator_function <- c(indicator_function, 1) 
      sum_in_set = sum_in_set + abs(data$values[i])
    } else{
      indicator_function <- c(indicator_function, 0) 
    }
  }
  
  data["indicator"] <- indicator_function # Add vector as a new column of the dataframe
  
  # Get the counts of overlapping genes
  # hit <- sum(with(data, indicator == 1)) 
  # misses <- sum(with(data, indicator == 0)) 
  
  # Create a list to allow multiple returns
  my_list <- list("data" = data, "sum_in_set" = sum_in_set)
  
  return(my_list)
} # Step 1

# Takes the dataframe and the sum of the overlapping genes
# Returns the updated dataframe and the MES rank 
computing_sk_ES <- function(data, sum_in_set){
  
  misses <- sum(with(data, indicator == 0))
  
  running_ES <- c()
  
  ES = 0
  MES_value = 0 
  
  for(i in data$ranking){ 
    if(data$indicator[i] == 1){
      ES = ES + (abs(data$values[i]) / sum_in_set)
      running_ES <- c(running_ES, ES)
    } else {
      ES = ES - (1 / misses)
      running_ES <- c(running_ES, ES)
    }
    if(abs(ES) > abs(MES_value)){ 
      MES_value = ES
    }
  }
  
  data["running_ES"] <- running_ES 
  
  MES_rank <- which(data$running_ES == MES_value) # To know the ranking of the MES
  
  my_list <- list("data" = data, "MES_rank" = MES_rank, "MES_value" = MES_value)
  
  return(my_list)
} # Step 2 and 3

# Takes as input the dataframe and the number of permutations to be completed
# Returns a vector with the MES of each permutation
phenotype_permutation <- function(data_permutation, permutation){
  
  # Create a copy of the dataframe that contains the names, values and the indicator
  data_permuted <- data_permutation[,1:2]
  data_permuted['indicator'] <- data_permutation[,4]
  
  ES_null <- c() # Vector to keep the MES computed in each permutation
  pos_null <- c()
  neg_null <- c()
  
  for(i in 1:permutation){ # (5.3)
    
    # We make the permutation over the values
    # In this way we do not need to assess the overlaping
    data_permuted$values <- sample(data_permuted$values) # The sample() function allows to perform random permutations 
    
    # We need to get the absolute sum of the rank metric for those genes that overlap
    permuted_sum_in_set <- sum(abs(data_permuted[which(data_permuted$indicator == 1), 2]))
    
    # We directly make the ranking of the new dataframe
    data_permuted <- data_permuted[order(data_permuted$values),] # Ordering the data frame by the value of expression
    data_permuted["ranking"] <- c(length(data_permuted$names):1)
    data_permuted <- data_permuted[order(data_permuted$ranking),] # Order the data frame by the ranking order
    
    misses <- sum(with(data_permuted, indicator == 0))
    
    ES = 0
    MES_value = 0 
    
    for(i in data_permuted$ranking){ 
      if(data_permuted$indicator[i] == 1){
        ES = ES + abs(data_permuted$values[i]) / permuted_sum_in_set
      } else {
        ES = ES - (1 / misses)
      }
      if(abs(ES) > abs(MES_value)){ 
        MES_value = ES
      }
    }
    
    ES_null <- c(ES_null, MES_value)
    
    if(MES_value > 0){
      pos_null <- c(pos_null, MES_value)
    } else{
      neg_null <- c(neg_null, MES_value)
    }
  }
  
  mylist <- list("ES_null" = ES_null, "pos_null" = pos_null, "neg_null" = neg_null)
  
  return(mylist)
} # Step 5.1

# Main function call
# Returns the final dataframe, the MES rank (order) and the MES value (score)
OurGSEA_plus <- function(data, current_candidate){
  
  overlap <- assessing_overlaping(data, current_candidate) # Step 1
  
  data_sk_ES <- computing_sk_ES(overlap$data, overlap$sum_in_set) # Step 2 and 3
  
  mylist <- list("data" = data_sk_ES$data, "MES_rank" = data_sk_ES$MES_rank, "MES_value" = data_sk_ES$MES_value)

  return(mylist)
}
```

## 1. Pipeline
**STEP 0: Processing the input files**
Initially a pre-ranked list and a gene list are used as inputs. The pre-ranked files are necessarily loaded by the user in .rnk format. In the case of the gene list, the user can use its own gene list, loaded in .tsv, .csv or .txt, or select one of the available (Hallmark collection). To assure a correct execution both inputs are pre-processed. Particularly, the GSEA+ application manages multiple input comparisons, therefore more than one pre-ranked list and gene set can be uploaded to be executed.

**STEP 1: Enrichment Score (ES) computation.**
To compute the ES it is necessary to firstly assess the overlap between genes in both inputs. In this intermediate step we attempt to compare the input files by defining two sets: genes in both lists (hit) and genes not present in the gene list (misses). To do so, we traverse all the genes in the pre-ranked gene set while updating a binomial variable that tracks the overlapping between inputs: 1 if present and 0 if absent. 
To correctly compute the Enrichment Score (ES) we need to consider, as stated by Subramanian et al., that the magnitude of the increment depends on the correlation of the gene with the phenotype.
We use a running sum to keep track of the ES computations. In this sense, the ES value will increase for those genes overlapping and decrease otherwise, allowing to draw the enrichment curve. The value that we subtract is fixed, and is defined as one divided by the number of genes that do not overlap. However, the addition score is dependent on the expression value of each of the overlapping genes: it corresponds to the absolute value of the differential expression of a gene divided by the sum of the expression values of all the overlapping genes.

**STEP 2: Drawing the GSEA plot.**
The classical GSEA plot is formed by three panels. The top panel corresponds to the enrichment profile, and the shape of the ES relates to the arrangement of the genes in the gene set compared with the whole expression set. The higher absolute ES matches the maximum ES (MES) and allows to delimit the leading-edge subset. The middle panel shows where the gene set members appear in the ranked gene list. The bottom panel represents the ranked list metric in a curve, which measures the gene’s correlation with a phenotype.

**STEP 3: Significance testing.**
GSEA employs permutation methods (= re-sampling) to generate a null distribution for each gene set. We used a phenotype permutation to randomly swap the sample labels and re-calculate each time the ES. With such approach we obtain how widely the ES varies and how often two groups are effectively the same. Therefore, step 1 is completed as many times as number of permutations. In each iteration the MES of the sampled data is stored to create a plot of the corresponding ES null distribution. 
From the null distribution the p-values is computed. The p-value is the probability of observing a statistic more extreme under the null hypothesis. To account for multiple hypotheses the significance level is adjusted by normalizing the ES (NES) for the gene set size. A positive NES represents that the genes in the gene set are mostly placed at the top of the pre-ranked list, and the opposite situation for a negative NES. The proportion of false positives is further controlled by calculating the false discovery rate (FDR) q-value. 

## 2. Exploring the input data
In order to run a GSEA analysis a preranked list and a gene set need to be selected. For this execution a total of `params$num_prerank` pre-ranked lists and `params$num_geneset` gene set were selected to run GSEA. 

Loaded files:
Pre-ranked lists: `params$names_prerank`
Gene sets: `params$names_geneset`

## 3. Exploring the outputs
This section will display the classical GSEA plots and their summary tables associated. Both results will be displayed for all the possible comparisons between the selected input files (all-to-all comparisons). The null ES distribution plot will be also incorporated to assess the significance. The actual (also called observed) MES (of each scenario) is also displayed in the plot.

### 3.1 Classical GSEA charts
A total of `params$num_plots` were generated. Each comparison is labeled with a number. This number will allow us to identificate the results of each comparison.

```{r plots, echo=FALSE}
count = 0

for(data_frame in params$gene_list){ # Traverse all the input gene lists
  for(list in 1:length(params$candidate_list)){ # Traverse all the candidate gene lists
    
    current_candidate <- c(params$candidate_list[list]) # Select a candidate gene list
    current_candidate <- current_candidate[[1[1]]] # Change the format
    
    GSEA <- OurGSEA_plus(data_frame, current_candidate) # Call to perform all GSEA functions
    
    count = count + 1
    id <- paste("Comparison: ", count)
    
    ########################
    # Drawing the GSEA plot
    ########################
    
    min <- min(GSEA$data$running_ES)
    max <- max(GSEA$data$running_ES)
            
    GSEA_plot <- ggplot() +
            geom_line(data = GSEA$data, mapping = aes(x = ranking, y = running_ES), color = params$color, size = params$curve_size) + 
            theme(
              panel.background = element_rect(fill = params$color_panel),
              plot.background = element_rect(fill = params$color_plot),
              plot.title = element_text(color = params$color_title, size = params$title_size, face = "bold", hjust = params$title_position)) +
            labs(x = "Rank in Ordered Dataset", y = "Running Enrichment Score", color = "Comparison Number") 

      # Add MES line
      if(params$add_mes == 1){
        GSEA_plot <- GSEA_plot + geom_vline(mapping = aes(xintercept = GSEA$MES_rank), linetype = "dashed", color = params$color_dashed)
      }
        
      # Add panel 2 (barcode + color range)
      if(params$add_panel2 == 1){
        up_colors <- colorRampPalette(c("red", "#FFFFFF")) # Ranges from red to white
        down_colors <- colorRampPalette(c("#FFFFFF", "#1b98e0")) # Ranges from white to blue
        
        hit_rank <- data.frame(which(GSEA$data$indicator == 1))
        indicator1 <- which(GSEA$data$indicator == 1)
        
        min <- min(GSEA$data$running_ES)
        max <- max(GSEA$data$running_ES)
        
        pos_val <- sum(GSEA$data$values > 0)
        if(pos_val > 0){
          limit <- which(GSEA$data$values == min(GSEA$data$values[GSEA$data$values > 0]))
        } else{
          limit <- 0
        }
        
        GSEA_plot <- GSEA_plot + geom_rug(data = hit_rank, mapping = aes(x = indicator1), 
                                          sides = "b", color = params$color_panel2, length = unit(0.1, "npc")) + ylim(min - 0.1, max) + 
          geom_rug(data = GSEA$data, mapping = aes(x = ranking), color = c(up_colors(limit), down_colors(nrow(GSEA$data) - limit))) +
          labs(x = NULL)
        
        # Highlight line when searched gene is overlapping
        if(is.null(params$selection_genes) == FALSE){
          filtered_data <- GSEA$data %>% filter(names %in% genes)
          filtered_data <- filtered_data %>% filter(indicator == 1)
          
          # Draws a line to represent the position of the gene (if the gene is present, indicator = 1)
          GSEA_plot <- GSEA_plot + geom_rug(data = filtered_data, mapping = aes(x = ranking), 
                                            sides = "b", length = unit(0.15, "npc"), color = params$color_search_gene, size = 2)
        }
      }
  
      # Display panel 3
      if(params$add_panel3 == 1){
        GSEA_plot <- GSEA_plot +
          theme(
            axis.ticks.x = element_blank(),
            axis.text.x = element_blank())
        
        threshold <- max(abs(min(GSEA$data$values)), abs(max(GSEA$data$values)))
        
        p2 <- ggplot(data = GSEA$data, mapping = aes(x = ranking, y = values)) + geom_col() + 
          labs(caption = id, y = "Ranked List Metric", x = "Rank in Ordered Dataset") +
          theme_classic() + ylim(-threshold, threshold)
        
        if(params$line_exp == 1){
          p2 <- p2 + geom_vline(mapping = aes(xintercept = limit[1]), linetype = "dashed", color = params$color_exp_line) 
        }
        
        if(params$red_blue == 2){
          p2 <- p2 + geom_col(color = params$color_exp)
        } else{
          p2 <- p2 + geom_col(color = ifelse(GSEA$data$values > 0, 'red', 'blue'))
        }
        
        GSEA_plot <- plot_grid(GSEA_plot, p2, ncol = 1, nrow = 2, rel_heights = c(2,1))
        
      } else{
        GSEA_plot <- GSEA_plot + labs(x = "Rank in Ordered Dataset")
      }
    # plots_list <- list(plots_list, GSEA_plot) # Save all the plots in a single list
    print(GSEA_plot)
  }
}
```

### 3.2. Summary table
The more important metrics and statistics have been collected in the table below. Each row presents the data of a particular comparison. The columns' guidelines are the following:

* Col. 1: Displays the label used to identificate each of the comparisons between a pre-ranked list and a gene set. It is linked to the legend of the GSEA plot.
* Col. 2|3: Contains the name of the pre-ranked lists and the gene sets.
* Col. 4|5: Total sum of genes in each of the pre-ranked lists and gene sets.
* Col. 6: Indicates the up-regulated class. Positive if enrichment in the left, and negative otherwise.
* Col. 7: Gives the values of the Maximum Enrichment Score (MES). The MES value indicates the maximum deviance of the ES from 0.
* Col. 8: Displays the rank were the MES is located.")
* Col. 9: Total sum of genes that are present in both input files. Correspond to the number of lines vertically displayed in the plot.
* Col. 10: Total sum of genes that form the leading-edge subset. The leading-edge subset is defined by the MES. If positive enrichment we will consider those genes before the MES, and if negative enrichment the genes after the MES will be considered.
       
```{r summary_table, echo=FALSE}
params$important_values[c(1,2,3)]

params$important_values[c(1,4,5,6,7,8,9)]
```

### 3.3. Significance estimation
To evaluate the significance a permutation method is completed iterating 1000 times. The maximum ES (MES) of each iteration is stored and displayed in a histogram. Additionally, the observed MES (i.e. the MES computed from the original data without randomization) is also displayed as a dashed line. The null ES distributions of each of the permutations are listed below.

```{r null_plots, echo=FALSE}
count = 0
null_dist <- c()
permutation_num = 1000 # Defining the number of times we will make the random sampling

for(data_frame in params$gene_list){ # Traverse all the input gene lists
  for(list in 1:length(params$candidate_list)){ # Traverse all the candidate gene lists
    
    current_candidate <- c(params$candidate_list[list]) # Select a candidate gene list
    current_candidate <- current_candidate[[1[1]]] # Change the format
    
    GSEA <- OurGSEA_plus(data_frame, current_candidate) # Call to perform all GSEA functions

    count = count + 1
    id <- paste("Comparison: ", count)

    ################################
    # Drawing the NULL distribution
    ################################
    
    ES_nulldist <- phenotype_permutation(GSEA$data, permutation_num) # Call to perform the phenotype permutation
    null_dist <- c(null_dist, list(ES_nulldist))
    
    permutation <- ES_nulldist$ES_null
    null <- data.frame(permutation)
    
    NULL_plot <- ggplot(data = null, mapping = aes(x = permutation)) + geom_histogram(bins = 30) + theme_classic() +
      geom_vline(data = GSEA$data, mapping = aes(xintercept = GSEA$MES_value), linetype = "dashed", color = params$color_dashed1) +
      labs(x = "ES", y = "P(ES)", title = "Random ES Distribution", caption = id) +
      theme(
        panel.background = element_rect(fill = params$color_panel1),
        plot.background = element_rect(fill = params$color_plot1),
        plot.title = element_text(color = params$color_title1, face = "bold", hjust = 0.5))
    
    print(NULL_plot)

  }
}
```

To numerically assess the significance of the obtained results the P-values, NES, FDR (q-value) and the TAGS, LIST and SIGNAL percentages were computed and collected in the following table.

```{r summary_stats, echo=FALSE}
params$important_statistics
```